A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-02-24 manuscript.tex
:memo: cluster interpretation txt moved to Rmd

Performance evaluation often means assessing the accuracy, stability, and separation or purity of the clustering \citep{keogh2003}. Importantly, any evaluation of the results depends on the research questions, the data, and the methods used. However, broadly speaking, evaluation methods can be categorized based on whether the true cluster labels are known or not \citep{saxena2017}. If true class labels are known, the cluster assignments can be compared to the true class labels --- using measures such as the F-measure, adjusted Rand index, mutual information, and normalized mutual information \citep[i.e., external evaluation; e.g.,][]{liao2005}. However, if the true cluster assignments are unknown, as with our psychological time series, the quality of the clusters is assessed based on the characteristics of the data itself \citep[i.e., internal evaluation; e.g.,][]{Aghabozorgi2015}.

@@

The interpretation of feature-based time series clustering in psychology involves understanding the meaning and implications of the clusters obtained from the analysis. Here, it is also important to consider the interpretability and practical significance of the clusters. While clusters with a high degree of internal cohesion may be more statistically robust, they may not necessarily be meaningful or useful in the context of the specific research question being addressed \citep{kaufman1990}. To gain a qualitative understanding of the extracted clusters users can use both the patterns that emerge from the clusters themselves as well as comparisons to between-person differences not included in the initial clustering \citep[e.g.,][]{monden2022}. As for the inspection of the clusters themselves, one common approach is comparing the clusters based on the average values of meaningful features \citep{Kennedy2021}. Assessing the clusters based on their average features allows users to gain an understanding of the developmental profiles. One cluster might, for example, have a relatively strong linear increase in well-being, or participants in another cluster might have a relatively high variability of negative emotions. Another common option for interpretation is finding an appropriate prototype to represent the cluster's trajectory. This is often easiest with embedded clustering methods because in many cases a cluster-level model is estimated as part of the expectation–maximization procedure \citep[e.g.,][]{denteuling2021}. For medoid-based clustering algorithms, a common approach is simply using cluster medoid as the prototype \citep{kaufman1990}. For most other cases, it is often recommended to use some form of average over time of the responses within the cluster \citep{niennattrakul2007}\footnote{It is important to note, however, that direct comparability can be a concern, and often times some subset selection or nonlinear alignment is necessary \citep[e.g.,][]{gupta1996}.}. These prototypes can then, for example, be inspected visually to identify qualitative differences between the cluster prototypes. An additional option for interpretations are out-of-cluster comparisons, where users can assess whether the clusters differ on some stable between-person differences, such as personality, age, or situational circumstances \citep[e.g.,][]{monden2022}.

\input{latex-results/06ClusterEvaluationInterpretation.tex}

>>>>

\subsubsection{Performance}
Performance evaluation often means assessing the accuracy, stability, and separation or purity of the clustering \citep{keogh2003}. Importantly, any evaluation of the results depends on the research questions, the data, and the methods used. However, broadly speaking, evaluation methods can be categorized based on whether the true cluster labels are known or not \citep{saxena2017}. If true class labels are known, the cluster assignments can be compared to the true class labels --- using measures such as the F-measure, adjusted Rand index, mutual information, and normalized mutual information \citep[i.e., external evaluation; e.g.,][]{liao2005}. However, if the true cluster assignments are unknown, as with our psychological time series, the quality of the clusters is assessed based on the characteristics of the data itself \citep[i.e., internal evaluation; e.g.,][]{Aghabozorgi2015}.

@@

\subsubsection{Interpretation}
The interpretation of feature-based time series clustering in psychology involves understanding the meaning and implications of the obtained clusters. In order to make sense of the clustering results, we here focus on three general aspects of the results \citep{kaufman1990}. (1) Assessing differences between the clusters in the original time series features, (2) comparing the clusters based on prototype developments, (3) comparing the clusters based on between-person differences that were not included in the initial clustering.

\input{latex-results/07ClusterEvaluationInterpretation.tex}
